---
title: 'Fact sheet: Epidemiologist'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

## Your role
As the epidemiologist on the team, your job is to help your team make informed decisions on the target population and study design considerations, including: 

- Select appropriate regions and health facilities.
- Decide on the number of clusters (health facilities) to include.
- Use expected malaria cases per month and test positivity rates to plan enrollment.

Navigate back to the [group activity instructions](module6_scenarios.Rmd) if you need to!

#### **Malaria prevalence, incidence and test positivity rate**

*Use the buttons to download as .csv or .xls if you prefer*
```{r echo=F}
epi <- data.frame(
  region = c("Coastal Region", "Highland Region", "Forest Region", "Urban Region"),
  prevalence = c("12%", "5%", "28%", "8%"),
  incidence_per_1000_per_year = c(40, 15, 90, 25),
  test_positivity_rate = c("18%", "7%", "35%", "10%"))

datatable(epi, 
          extensions = 'Buttons',
          options = list(
            dom = 'Bfrtip',
            buttons = c('csv', 'excel'),
            searching = FALSE
          ),
          rownames = FALSE
)
```
