---
title: 'Activity 5: Dealing with overdispersion in multi-cluster studies'
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# options(repos = c(
#   plasmogenepi = "https://plasmogenepi.r-universe.dev",
#   CRAN = "https://cloud.r-project.org"
# ))

library(learnr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(DRpower)
knitr::opts_chunk$set(warning = FALSE,
                      echo = FALSE)

# needed for longer calculations (eg DRpower) note - in seconds
tutorial_options(exercise.timelimit = 120)

set.seed(3)

# mock up pfhrp2/3 deletion data
dat <- data.frame(Site = c("Chalinze", "Ibwaga", "Lukali", "Malolo", "Mafene", "Mpendo", "Nhinhi", "Rudi")) |>
  mutate(Confirmed_malaria = c(100, 80, 100, 100, 95, 100, 50, 100),
         pfhrp2_deleted = DRpower:::rbbinom_reparam(n_clust = 8, N = Confirmed_malaria, p = 0.2, rho = 0.1),
         prevalence = round(pfhrp2_deleted / Confirmed_malaria, 3))

# save variables
n <- dat$Confirmed_malaria
x <- dat$pfhrp2_deleted
p <- dat$prevalence
```

## Introduction

Welcome to Activity 5: **Dealing with overdispersion in multi-cluster studies**

In this activity, we will explore the unique features of multi-cluster studies that differentiate them from single-cluster studies. In malaria surveillance, it is common to pool information from multiple sites, such as health facilities, towns, or even regions within a country. Combining data across sites allows us to draw broader conclusions at a higher geographic or administrative level. However, multi-cluster studies often face the challenge of *overdispersion*, where variability between sites exceeds what we would expect by chance. Overdispersion can reduce the precision of our estimates and lower statistical power, making it crucial to account for this phenomenon in our analyses.

### Learning Outcomes

By the end of this tutorial, you will be able to:

- Detect overdispersion
- Quantify overdispersion using three metrics; 1) the design effect, 2) the effective sample size, and 3) the intra-cluster correlation coefficient
- Understand the impact of overdispersion on statistical efficiency
- Construct confidence intervals and perform hypothesis tests that account for overdispersion

*Disclaimer: The scenarios in this document are entirely fictitious. While real place names are used, the data itself is artificial and designed for teaching purposes only. It does not necessarily represent the real epidemiological situation in these locations.*

## Analyzing data from a multi-cluster *pfhrp2/3* deletion study

### Background

You are collaborating with the Tanzanian National Malaria Control Programme (NMCP) to investigate the prevalence of *pfhrp2/3* gene deletions in the Dodoma region of Tanzania. These gene deletions pose a significant threat to malaria control efforts because they can lead to parasites being undetectable by rapid diagnostic tests (RDTs) that rely on the HRP2 protein. Undetected cases may lead to delays in treatment or missed malaria diagnoses, undermining effective case management.

A multi-cluster study has been performed in 8 sites within the Dodoma region. The results of this study are shown below:

```{r}
# create table with kable
kable(dat, "html", col.names = c("Site", "Confirmed Malaria<br>(n)", "pfhrp2/3 Deleted<br>(x)", "pfhrp2/3 Deletion Prevalence<br>(p = x / n)"), escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(1, width = "3cm") |>
  column_spec(2, width = "3cm") |>
  column_spec(3, width = "3cm") |>
  column_spec(4, width = "4cm") |>
  footnote(general = "\nTable 1: pfhrp2/3 deletion data broken down by site", general_title = "")
```

In the R chunks below, you will be able to access the columns of this table through the variables `n`, `x`, and `p`, as shown here:

```{r sandbox-1, exercise=TRUE}
# you can access the following vectors
print(n)
print(x)
print(p)
```


### Estimating the global prevalence

We want to use the information over all 8 sites to estimate the prevalence of *pfhrp2/3* deletions in the Dodoma region as a whole. This is often refered to as the "global" prevalence estimate. We will use $\hat{p}$ to denote the global prevalence, and $\hat{p}_i$ to denote the site-level prevalence in the $i^\text{th}$ site.

A common way to calculate $\hat{p}$ is to take the mean over sites.

$$
\hat{p} = \frac{1}{c} \sum_{i=1}^c \hat{p}_i
$$

where there are $c$ sites. In our case $c = 8$.

Complete the following R code to calculate `p_global` as the mean prevalence over sites. Remember that the values in **Table 1** are available through the variables `n`, `x` and `p`.

```{r global-1, exercise=TRUE}
# calculate p_global as mean prevalence over sites
p_global <- 

print(p_global)
```

```{r global-1-solution}
# calculate p_global as mean prevalence over sites
p_global <- mean(p)

print(p_global)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that $\hat{p} = 0.172$, or 17.2% prevalence.

</details>
</br>

This calculation ignores differences in sample sizes between sites. For example, the Nhinhi site is given just as much weight as the Rudi site, despite having half the number of confirmed malaria cases. A different approach is to calculate $\hat{p}$ as the *weighted* average of the site-level prevalence, where the weights are given by the sample sizes. This is mathematically equivalent to summing the numerator, summing the denominator, and then taking the ratio:

$$
\hat{p} = \frac{\sum_{i=1}^c x_i}{\sum_{i=1}^c n_i}
$$

Complete the following R code to calculate `p_global` as the *weighted* mean prevalence over sites:

```{r global-2, exercise=TRUE}
# calculate p_global as weighted mean prevalence over sites
p_global <- 

print(p_global)
```

```{r global-2-solution}
# calculate p_global as weighted mean prevalence over sites
p_global <- sum(x) / sum(n)

print(p_global)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that $\hat{p} = 0.183$, or 18.3% prevalence.

</details>
</br>

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  Why does the weighted mean give a higher prevalence in this example than the unweighted mean? Look at the prevalence and sample size columns in **Table 1** for a clue.
  </span>
</div>

Although we have seen two ways of calculating the global prevalence, neither approach is more correct than the other, they just have different strengths and weaknesses. The unweighted mean treats each site as a single observation, which is a robust approach when there may be overdispersion. On the other hand, the weighted mean avoids small clusters having a large influence on the final estimate. For the purposes of this activity, we will use the unweighted mean as our global estimate.

### Detecting overdispersion

Now that we have an estimate of the global prevalence, we can look for *overdispersion* in the data. If all patients enrolled in the study were completely independent, meaning they all had the same probability $\hat{p}$ of carrying the *pfhrp2/3* deletion, then we would expect to see a certain level of variation between sites. Most of the time, the site-level prevalence should be within the following 95% interval:
$$
\hat{p} \pm z_{1 - \alpha/2}\sqrt{\frac{\hat{p}(1 - \hat{p})}{n_i}}
$$
where $n_i$ is the sample size in the $i^{\text{th}}$ site. This is a 95% interval, meaning we should expect site-level prevalence to fall within this range around 95% of the time. For our study involving 8 sites, we *may* see one value outside this range by chance, but it would be very unusual to see more than one value outside this range.

The plot below shows the site-level prevalence in red. The global mean prevalence of 17.2% is shown as a horizontal dashed line, and the 95% interval is shown as an error bar.

```{r, echo=FALSE, fig.width=8, fig.height=5}
p_global <- mean(p)
z <- qnorm(1 - 0.05 / 2)
MOE <- z*sqrt(p_global*(1 - p_global) / n)

df_errorbar <- data.frame(Site = dat$Site,
                          p = p,
                          MOE = MOE)

dat |>
  mutate(non_deleted = 1 - prevalence) |>
  select(Site, prevalence, non_deleted) |>
  pivot_longer(cols = c(prevalence, non_deleted)) |>
  ggplot() + theme_bw() +
  geom_bar(aes(x = Site, y = value, fill = name), stat = "identity") +
  geom_errorbar(aes(x = Site, ymin = p_global - MOE, ymax = p_global + MOE), width = 0.2, data = df_errorbar) +
  geom_hline(yintercept = p_global, linetype = "dashed") +
  scale_fill_manual(values = c("grey", "firebrick1"), labels = c("non-deleted", "deleted"), name = NULL) +
  ylab("Prevalence")
```

```{r quiz-overdispersion-1}
quiz(caption = "QUIZ - Detecting overdispersion",
  question("How many sites have a prevalence outside the expected 95% interval?",
           allow_retry = TRUE,
           answer("0", correct = FALSE),
           answer("2", correct = FALSE),
           answer("4", correct = FALSE),
           answer("6", correct = TRUE), 
           correct = "That is correct! There are six sites that have prevalence either above or below the 95% confidence interval.", 
           incorrect = "Hint: a site-level prevalence is outside the range if it is either above the upper error bar, or below the lower error bar."
  ),
  
  question_radio("From this plot, does the prevalence appear to be over-dispersed, under-dispersed, or neither?",
           allow_retry = TRUE,
           answer("Over-dispersed", correct = TRUE),
           answer("Under-dispersed", correct = FALSE),
           answer("Neither", correct = FALSE),
           correct = "That is correct! The presence of this many sites outside the expected range is unlikely to happen by chance, and is an indication of overdispersion.", 
           incorrect = "Hint: 'dispersion' describes the spread of data. Is there more or less spread than we would expect?."
  ),
  
  question_checkbox("What could cause this overdispersion? Select all correct responses.",
           allow_retry = TRUE,
           answer("Limited migration and gene flow between sites.", correct = TRUE),
           answer("Variation in treatment practices leading to different selective pressures between sites.", correct = TRUE),
           answer("Local outbreaks within sites, driven by infected individuals carrying *pfhrp2/3* deletions.", correct = TRUE),
           correct = "That is correct! A wide range of factors can lead to overdispersion, meaning it is extremely common in multi-cluster studies.", 
           incorrect = "Hint: that is one explanation, but is it the only possible explanation?"
  ),
  
  question("Why does the Nhinhi site have a slightly wider 95% interval than the other sites?",
           allow_retry = TRUE,
           answer("The Nhinhi site has higher prevalence of *pfhrp2/3* deletions.", correct = FALSE),
           answer("The diagnostic methods used at the Nhinhi site are less accurate.", correct = FALSE),
           answer("The Nhinhi site is geographically isolated.", correct = FALSE),
           answer("The sample size in Nhinhi is lower than other sites.", correct = TRUE),
           correct = "That is correct! A smaller sample size will lead to a wider 95% interval.", 
           incorrect = "Hint: look at the formula for the 95% interval. Which factor in the formula means it will vary between sites?"
  )
)
```


## Quantifying overdispersion

### The design effect 

One way of quantifying the effect of overdispersion is through the **design effect**, denoted $D_{\text{eff}}$. We can estimate the design effect by calculating the observed variance between sites and dividing this by the variance that we would expect under simple random sampling (SRS). If the data are over-dispersed then the observed variance will be greater than expected, meaning $D_{\text{eff}}$ will be greater than 1.

The observed variance, or sample variance, can be calculated using the formula:

$$
\text{Var}_{\text{obs}} = \frac{1}{c-1}\sum_{i=1}^c (\hat{p}_i - \hat{p})^2
$$
This formula is found in many areas of statistics. We don't need to calculate this value by hand, instead we can do it very easily in R using the `var()` function:

```{r, echo=TRUE}
# calculate observed variance between sites
var_observed <- var(p)

print(var_observed)
```

Next, we need to calculate the variance that we would expect under simple random sampling (SRS). This is given by:

$$
\text{Var}_{\text{SRS}} = \frac{1}{c} \sum_{i=1}^c \frac{\hat{p}(1 - \hat{p})}{n_i}
$$
We can calculate this in R as follows:

```{r, echo=TRUE}
# calculate expected variance under SRS
var_SRS <- mean(p_global*(1 - p_global) / n)

print(var_SRS)
```

Finally, we can calculate the design effect as the ratio of these two quantities:
$$
D_{\text{eff}} = \frac{\text{Var}_{\text{obs}}}{\text{Var}_{\text{SRS}}}
$$
Complete the following R code to calculate the design effect:

```{r deff-params}
p_global <- mean(p)
var_observed <- var(p)
var_SRS <- mean(p_global*(1 - p_global) / n)
```

```{r deff-1, exercise=TRUE, exercise.setup = "deff-params"}
# calculate design effect from the variances
D_eff <- 

print(D_eff)
```

```{r deff-1-solution}
# calculate design effect from the variances
D_eff <- var_observed / var_SRS

print(D_eff)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that $D_{\text{eff}} = 12.79$.

</details>
</br>

The design effect is a measure of the statistical *inefficiency* of a study design. Larger values indicate less efficient designs, with a value of $D_{\text{eff}}=1$ meaning the study design is as efficient as SRS, which is seen as the gold standard for statistical efficiency. Our observed value of $D_{\text{eff}}=12.79$ indicates that our study has a high level of statistical inefficiency, which is caused by overdispersion in the data.

```{r quiz-deff-1}
quiz(caption = "QUIZ - Interpreting the design effect",
  question_checkbox("A high value of the design effect means we can expect to see (check all correct responses):",
           allow_retry = TRUE,
           answer("Less precise estimates of the global prevalence $\\hat{p}$", correct = TRUE),
           answer("Lower power", correct = TRUE),
           answer("Larger sample sizes needed", correct = TRUE),
           correct = "That is correct! A high design effect means we have a statistically inefficient design, which will impact our precision, power, and sample size.", 
           incorrect = "Hint: that is a correct response, but is it the only correct response?"
  )
)
```

### The effective sample size

Another way to quantify the impact of overdispersion is through the *effective sample size*, $N_{\text{eff}}$. The effective sample size tells us how many perfectly independent samples would be needed in order to achieve the same level of statistical efficiency as present in our design. In other words, it tells us how large our study would need to be if we could get rid of overdispersion completely. We calculate $N_{\text{eff}}$ by dividing the true total sample size by the design effect:

$$
N_{\text{eff}} = \frac{\sum_{i=1}^c n_i}{D_{\text{eff}}}
$$

Complete the following R code to calculate an effective sample size:

```{r neff-params}
p_global <- mean(p)
var_observed <- var(p)
var_SRS <- mean(p_global*(1 - p_global) / n)
D_eff <- var_observed / var_SRS
```

```{r neff-1, exercise=TRUE, exercise.setup = "neff-params"}
# calculate effective sample size based on total sample size and design effect
N_eff <- 

print(N_eff)
```

```{r neff-1-solution}
# calculate effective sample size based on total sample size and design effect
N_eff <- sum(n) / D_eff

print(N_eff)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that $N_{\text{eff}} = 56.7$.

</details>
</br>

Our effective sample size is just 56.7, despite a total sample size of 725. This means that enrolling only 57 patients through true random sampling from the population would have provided the same amount of information as our multi-cluster study with 725 participants. However, this doesn’t imply that a simple random sampling approach would have been better—such an approach would likely have been less feasible and more expensive. Instead, the concept of effective sample size gives us a clearer understanding of how overdispersion leads to a loss of information, helping us better account for its impact in multi-cluster studies.


### The intra-cluster correlation coefficient

The third and final measure that we will consider is the *intra-cluster correlation coefficient* (ICC), denoted $r$. This is a measure between 0 and 1 that describes how correlated observations are *within* a cluster (site). If there is correlation with clusters then there *must* be overdispersion between clusters, and likewise if there is overdispersion then there *must* be intra-cluster correlation greater than 0.

The relationship between the ICC and the design effect is:

$$
D_{\text{eff}} = 1 + r(\bar{n} - 1)
$$
where $\bar{n}$ is the mean sample size over sites. Notice that when $r$ is zero the design effect equals 1, indicating that our study is as statistically efficient as simple random sampling. The larger the value of $r$, the higher the design effect and the more inefficient our design.

We can flip this equation around to give us the ICC as a function of the design effect:

$$
r = \frac{D_{\text{eff}} - 1}{\bar{n} - 1}
$$

Complete the following R code to calculate the value of the ICC from the design effect that we estimated above:

```{r ICC-1, exercise=TRUE, exercise.setup = "neff-params"}
# calculate the mean sample size
n_bar <- 

# calculate the ICC
ICC <- 

print(ICC)
```

```{r ICC-1-solution}
# calculate the mean sample size
n_bar <- mean(n)

# calculate the ICC
ICC <- (D_eff - 1) / (n_bar - 1)

print(ICC)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that $r = 0.132$.

</details>
</br>

The ICC can be harder to interpret than other measures like the design effect or effective sample size, but it does have certain advantages. We can think of the ICC as an *intrinsic* property of the population, while the design effect is only a measure of the inefficiency of *our study*. While $D_{\text{eff}}$ is influenced by $r$, it is also influenced by many other factors, most notably the sample size. If we were to double the sample size of a study then our design effect would change, but the ICC would stay exactly the same. This makes it more straightforward to compare values of $r$ between studies, while we have to take great care when comparing values of $D_{\text{eff}}$.

## Accounting for overdispersion in analysis

We have seen what overdispersion looks like in prevalence data, and how it can be quantified. Now we will learn how to incorporate overdispersion into our design and analysis phases.

### Overdispersion and confidence intervals

Let's look again at the *pfhrp2/3* data from Dodoma region:

```{r}
# create table with kable
kable(dat, "html", col.names = c("Site", "Confirmed Malaria<br>(n)", "pfhrp2/3 Deleted<br>(x)", "pfhrp2/3 Deletion Prevalence<br>(p)"), escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) |>
  column_spec(1, width = "3cm") |>
  column_spec(2, width = "3cm") |>
  column_spec(3, width = "3cm") |>
  column_spec(4, width = "4cm") |>
  footnote(general = "\nTable 2: pfhrp2/3 deletion data broken down by site", general_title = "")
```

We already estimated the prevalence in the region as the mean over sites, giving $\hat{p} = 0.172$. Now we want to construct a 95% CI around this estimate. For this, we need a modified version of the Wald confidence interval:

$$
\hat{p} \pm z_{1 - \alpha/2}\sqrt{\frac{\hat{p}(1 - \hat{p})}{N} D_{\text{eff}}}
$$
Here, $N$ is the total sample size summed over all sites, which in our case is $N = 725$. The main difference from our previous version of the Wald confidence interval is the inclusion of the design effect, $D_{\text{eff}}$. Notice that $D_{\text{eff}}$ *stretches* the interval, meaning a higher design effect would lead to greater uncertainty.

Complete the following R code to calculate a 95% CI on the global prevalence using this new version of the Wald formula. Note that you still have access to previously computed variables in this code box, including `n`, `x`, `p`, `p_global` and `D_eff`:

```{r wald-params}
p_global <- mean(p)
var_observed <- var(p)
var_SRS <- mean(p_global*(1 - p_global) / n)
D_eff <- var_observed / var_SRS
```

```{r Wald-1, exercise=TRUE, exercise.setup = "wald-params"}
# get total sample size and z
N <- 
z <- 1.96

# calculate margin of error
m <- 

# calculate lower and upper limits
CI_lower <- 
CI_upper <- 

print(c(CI_lower, CI_upper))
```

```{r Wald-1-solution}
# get total sample size and z
N <- sum(n)
z <- 1.96

# calculate margin of error
m <- 1.96*sqrt(p_global*(1 - p_global) / N * D_eff)

# calculate lower and upper limits
CI_lower <- p_global - m
CI_upper <- p_global + m

print(c(CI_lower, CI_upper))
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that the CI ranges from 0.073 to 0.270, or in other words from 7.3% to 27.0%.

</details>
</br>

This CI is considerably wider than we would have obtained if we did not take overdispersion into account. The two confidence intervals are shown below:

```{r, fig.width=4, fig.height=4}
p_global <- mean(p)
var_observed <- var(p)
var_SRS <- mean(p_global*(1 - p_global) / n)
D_eff <- var_observed / var_SRS
N <- sum(n)

MOE1 <- 1.96*sqrt(p_global*(1 - p_global) / N)
MOE2 <- 1.96*sqrt(p_global*(1 - p_global) / N * D_eff)

rbind.data.frame(list(type = "Without\ndesign effect", y = p_global, ymin = p_global - MOE1, ymax = p_global + MOE1),
                 list(type = "With\ndesign effect", y = p_global, ymin = p_global - MOE2, ymax = p_global + MOE2)) |>
  ggplot() + theme_bw() +
  geom_errorbar(aes(x = type, ymin = ymin, ymax = ymax), width = 0.1) +
  geom_point(aes(x = type, y = y)) +
  scale_y_continuous(limits = c(0, 0.5), expand = c(0, 0)) +
  xlab("") + ylab("Global prevalence")
```

If we ignored overdispersion and assumed all observations were independent, we would risk overestimating the precision of our global prevalence estimate, leading to unwarranted over-confidence in its accuracy.

### Overdispersion and hypothesis testing

Similarly, when it comes to hypothesis testing, we have to factor overdispersion into our analysis. For example, imagine we want to compare our global estimate of the prevalence of *pfhrp2/3* deletions against a defined threshold of 10%. The appropriate statistical test here is the one-sample z-test for proportions, which we covered in an earlier module. However, we need to use a new formula for the test statistic ($Z$) that factors in the design effect:

$$
Z = \frac{\hat{p} - p_0}{\sqrt{\frac{\hat{p}(1 - \hat{p})}{N}D_{\text{eff}}}}
$$
Where $p_0$ is the threshold we are comparing against.

This can be calculated in R as follows:

```{r, echo=TRUE}
# define threshold
p0 <- 0.1

# calculate test statistic
Z <- (p_global - p0) / sqrt(p_global * (1 - p_global) / N * D_eff)
print(Z)
```

We obtain a test statistic of $Z = 1.43$. This does not exceed the critical values at $\pm1.96$, meaning we would fail to reject the null hypothesis. However, if we were to run the same analysis but ignoring the design effect, we would find $Z = 5.11$. This is highly significant, and would result in us rejecting the null hypothesis. This is an example where overdispersion is critically important, as it dictates the result of our hypothesis test! For something like a *pfhrp2/3* study, this could mean the difference between switching RDTs nationwide vs. sticking with the current brand of tests.

**Disclaimer: We do not advocate for the use of the z-test for the design of pfhrp2/3 deletion studies, or the analysis of the resulting data. We will learn about a much more powerful approach to this problem in the next module.**

## Accounting for overdispersion in design

### Overdispersion and power

Addressing overdispersion is just as critical in the design stage as it is in the analysis stage. For the one-sample z-test for proportions, we already saw the formula for power in a previous module. Now, we obtain a new formula that takes $D_{\text{eff}}$ into account:

$$
\text{Power} = 1 - \phi\left(z_{1-\alpha/2} - \frac{|p - p_0|}{\sqrt{\frac{p(1-p)}{N}}D_{\text{eff}}} \right)
$$
To make use of this formula, we have to assume known values for the true global prevalence $p$ and also the design effect $D_{\text{eff}}$. We also have to set the total sample size $N$ and the threshold $p_0$. We will assume $p = 0.2$, $D_{\text{eff}} = 10$, $N = 800$ and $p_0 = 0.1$.

We can calculate the power under this new formula in R as follows:

```{r, echo=TRUE}
# define assumed parameters
p <- 0.2
D_eff <- 10
N <- 800
p0 <- 0.1

# define z-values
z_alpha <- 1.96

# calculate expected value of test statistic
E_Z <- abs(p - p0) / sqrt(p * (1 - p) / N * D_eff)

# calculate power
power <- 1 - pnorm(z_alpha - E_Z)
print(power)
```

Under this study design we only have 61% power. Contrast this with power of 99.99998% that we obtain if we ignore the design effect! Clearly the inclusion of the design effect is severely hurting our power in this analysis, however, it is better to include the design effect and be realistic than to ignore the design effect and obtain results that are unreliable.

### Overdispersion and sample size

When it comes to the formula for minimum sample size under the z-test, we cannot assume a known value of the design effect. This is because the design effect itself depends on the sample size, so our reasoning becomes circular. However, in our equation for sample size we can assume a known value of the ICC as this is an intrinsic propertly of the population.

Our new minimum sample size formula becomes:

$$
n = \frac{1 - r}{\frac{c(p - p_0)^2}{(z_{1 - \alpha/2} + z_{1 - \beta})^2 p(1-p)} - r}
$$

Note that $n$ is the sample size *per cluster*, assuming there are $c$ clusters of equal size. We can calculate this in R as follows:

```{r, echo=TRUE}
# define assumed parameters
p <- 0.2        # prevalence of deletions
p0 <- 0.1        # threshold to compare against
r <- 0.05       # intra-cluster correlation
c <- 8          # number of clusters

# define z-values
z_alpha <- 1.96
z_beta <- qnorm(0.8)

# calculate minimum sample size
n <- (1 - r) / (c*(p - p0)^2 / ((z_alpha + z_beta)^2*p*(1 - p)) - r)

print(n)
```

Based on this analysis we would need 8 clusters of 70 samples. Compare this with the 8 clusters of just 16 samples we would need if we ignored intra-cluster correlation. Again, if we ignore this effect then we risk running a study that is too small, and ultimately does not have power to detect a real effect.
