---
title: 'Fact sheet: Health facility coordinator'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(DT)
```

## Your role
As the health facility coordinator on the team, your job is to help your team make informed decisions on the target health facilities (cluster) and study design considerations, including: 

- Select appropriate regions and health facilities.
- Decide on the number of clusters (health facilities) to include.
- Planning for the time it will take to enrol the total number of malaria-positive patients in each location.

Navigate back to the [group activity instructions](module6_scenarios.Rmd) if you need to!

**Health facility information**

*Use the buttons to download as .csv or .xls if you prefer*
```{r echo=F}
# Create a data frame for health facilities
health_facilities <- data.frame(
  region = c(
    rep("Coastal Region", 20),
    rep("Highland Region", 10),
    rep("Forest Region", 15),
    rep("Urban Region", 20)
  ),
  health_facility = c(
    paste0("HF", 1:20),    # Coastal Region
    paste0("HF", 21:30),   # Highland Region
    paste0("HF", 31:45),   # Forest Region
    paste0("HF", 46:65)    # Urban Region
  ),
  population_served = c(
    # Coastal Region Population Served
    12000, 14000, 16000, 18000, 20000, 22000, 24000, 26000, 28000, 30000,
    32000, 34000, 36000, 38000, 40000, 42000, 44000, 46000, 48000, 50000,
    # Highland Region Population Served
    10000, 12000, 14000, 16000, 18000, 20000, 22000, 24000, 26000, 28000,
    # Forest Region Population Served
    15000, 18000, 21000, 24000, 27000, 30000, 33000, 36000, 39000, 42000,
    45000, 48000, 51000, 54000, 57000,
    # Urban Region Population Served
    20000, 22000, 24000, 26000, 28000, 30000, 32000, 34000, 36000, 38000,
    40000, 42000, 44000, 46000, 48000, 50000, 52000, 54000, 56000, 58000
  ),
  expected_malaria_cases_per_month = c(
    # Coastal Region Expected Cases per Month
    40, 47, 53, 60, 67, 73, 80, 87, 93, 100,
    107, 113, 120, 127, 133, 140, 147, 153, 160, 167,
    # Highland Region Expected Cases per Month
    13, 15, 18, 20, 23, 25, 28, 30, 33, 35,
    # Forest Region Expected Cases per Month
    113, 135, 158, 180, 203, 225, 248, 270, 293, 315,
    338, 360, 383, 405, 428,
    # Urban Region Expected Cases per Month
    42, 46, 50, 54, 58, 63, 67, 71, 75, 79,
    83, 88, 92, 96, 100, 104, 108, 113, 117, 121
  )
)

# Render the datatable without the search bar
datatable(
  health_facilities,
  extensions = 'Buttons',
  options = list(
    dom = 'Bfrtip',
    buttons = c('csv', 'excel')
  ),
  rownames = FALSE
)
```
