---
title: "Activity 2: Sample Size Calculation Base on Margin of Error"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(dplyr)
library(ggplot2)
library(knitr)
knitr::opts_chunk$set(warning = FALSE,
                      echo = FALSE)

# needed for longer calculations (eg DRpower) note - in seconds
tutorial_options(exercise.timelimit = 120)
```

## Introduction

Welcome to Activity 2: **Sample Size Calculation Base on Margin of Error**

In this activity, we’ll explore how margin of error (MOE) arguments can guide sample size determination. This approach relies on the concept of *precision* - larger sample sizes lead to more precise estimates.

### Learning Outcomes

By the end of this tutorial, you will be able to:

- Calculate a 95% confidence interval from prevalence data.
- Calculate a minimum sample size using assumptions about prevalence and margin of error.
- Account for drop-out and malaria positive fraction.

*Disclaimer: The scenarios in this document are entirely fictitious. While real place names are used, the data itself is artificial and designed for teaching purposes only. It does not necessarily represent the real epidemiological situation in these locations.*

## Background

You have been recruited by the National Malaria Control Programme (NMCP) of the Democratic Republic of the Congo (DRC) to assist with study design. The NMCP is concerned about the potential spread of mutations that confer partial resistance to the drug combination Sulfadoxine-Pyrimethamine (SP). The *dhps* K540E mutation, which is known to be associated with high level SP resistance when found alongside other common mutations, has recently been found at 72% prevalence in neighbouring Uganda. In the last few weeks there have been anecdotal reports of SP failure in Rutshuru town, which lies in Eastern DRC close to the border with Uganda. The NMCP is concerned that drug resistant parasites may be flowing over the border.

The NMCP plans to conduct a cross-sectional study to estimate the prevalence of the *dhps* K540E mutation in Rutshuru town. **Your job is to work out the appropriate sample size for this study.**

## Results of a pilot study

Fortunately, a pilot study has already been conducted in Rutshuru town. This pilot study included 100 participants chosen at random from households within the town. These participants were tested for malaria via rapid diagnostic test (RDT). 23 people tested positive for malaria, and these samples were sent away for genetic sequencing. 19 samples were successfully sequenced, and 5 of these were found to carry the *dhps* K540E mutation.

```{r quiz-pilot1}
quiz(caption = "QUIZ - Estimating molecular marker prevalence",
     question("Which is the correct formula when calculating the prevalence of *dhps* K540E mutations from the pilot data?",
           allow_retry = TRUE,
           answer("23 / 100 = 23.0%", correct = FALSE),
           answer("19 / 100 = 19.0%", correct = FALSE),
           answer("5 / 23 = 21.7%", correct = FALSE),
           answer("5 / 19 = 26.3%", correct = TRUE), 
           correct = "That is correct! The numerator is the number of times we observed the K540E mutation, and the denominator is the number of *successfully sequenced* samples.", 
           incorrect = "Hint: the numerator should be the number of times we observed the K540E mutation, and the denominator should be the number of samples that had the *potential* to observe this mutation."
  )
)
```

Recall that we can use the following formula to calculate a 95% confidence interval on our prevalence estimate:

$$
\hat{p} \pm z_{1-\alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
$$
Complete the following R code to compute this interval:

```{r wald-1, exercise=TRUE}
# estimate the prevalence
x <- 
n <- 
p <- x / n

print(p)

# calculate the margin of error (MOE)
z_alpha <- 1.96
MOE <- z_alpha * sqrt(p*(1 - p) / n)

# compute lower and upper 95% limits
p - MOE
p + MOE
```

```{r wald-1-solution}
# estimate the prevalence
x <- 5
n <- 19
p <- x / n

print(p)

# calculate the margin of error (MOE)
z_alpha <- 1.96
MOE <- z_alpha * sqrt(p*(1 - p) / N)

# compute lower and upper 95% limits
CI_lower <- p - MOE
CI_upper <- p + MOE
c(CI_lower, CI_upper)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

Our point estimate is 26.3%. The 95% interval goes from 0.065 to 0.461, or in other words from 6.5% to 46.1%.

</details>
</br>

The 95% confidence interval is very wide, revealing there is considerable uncertainty about the true prevalence of *dhps* K540E mutations. While our best estimate is 26.3%, the plausible range spans from 6.5% to 46.1%, which the NMCP finds too broad to be practically useful. They therefore plan to conduct a follow-up study to obtain a more precise estimate.


## Calculating the appropriate sample size

When designing the new study, we will calculate the exact sample size needed to achieve a target margin of error (MOE). We can do this by rearranging the MOE formula to isolate the sample size ($n$) on the left side. If you’re comfortable with the math and would like to see the steps, follow along below. Otherwise, feel free to skip to Step 3 for the final formula.

**Step 1: Write down the formula for the MOE**

We will use the mathematical symbol $m$ for the MOE:

$$
m = z_{1-\alpha/2}\sqrt{\frac{p(1-p)}{n}}
$$
**Step 2: Square both sides**

$$
m^2 = z_{1-\alpha/2}^2 \frac{p(1-p)}{n}
$$
**Step 3: Multiply both sides by $n$ and divide both sides by $m^2$**

$$
n = z_{1-\alpha/2}^2 \frac{p(1-p)}{m^2}
$$

At the end of Step 3 we have a new formula that we can plug in values of $p$ and $m$ and it will tell us the appropriate sample size $n$.

Our reason for working through the derivation of this formula was to show how closely connected it is to the formula for the confidence interval. In fact, it is the same mathematical expression, just reverse engineered to be in terms of $n$!

Now we need to decide what values to assume for $p$ and $m$.

```{r quiz-prev}
quiz(caption = "QUIZ - Choosing assumed parameter values",
     question("Which of these are reasonable choices for $p$?",
           allow_retry = TRUE,
           answer("$p = 0.26$ as this was the prevalence found in the pilot data.", correct = FALSE),
           answer("$p = 0.72$ as this was the prevalence found in neighbouring Uganda", correct = FALSE), 
           answer("$p = 0.5$ as this is the most pessimistic assumption. It will lead to the largest sample size, and so by assuming this value we are robust to any other value of the prevalence.", correct = FALSE),
           answer("All of the above", correct = TRUE), 
           correct = "That is correct! Any of these methods are reasonable as long as we can justify them.",
           incorrect = "That is a reasonable choice, but it is not the only reasonable choice!"
  )
)
```

In this tutorial we will assume a value of $p=0.26$ to match the pilot data. The NMCP has decided that a MOE of 5% is acceptable. Complete the following R code to compute the resulting sample size:

```{r sample_size-1, exercise=TRUE}
# enter assumed values
p <- 
m <- 

# calculate the raw sample size
z <- 1.96
z^2*p*(1 - p) / m^2
```

```{r sample_size-1-solution}
# enter assumed values
p <- 0.26
m <- 0.05

# calculate the raw sample size
z <- 1.96
z^2*p*(1 - p) / m^2
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

We obtain a value of 295.65. We would round this *up* to $n=296$ to give a whole number for sampling.

</details>
</br>

The nice thing about sample size determination is that we can quickly confirm our calculation is correct:

**Optional exercise:** Try entering the values $p=0.26$ and $n=296$ into the 95% CI formula that we used in the pilot data analysis. If our calculations were correct, you should find that the resulting MOE is very close to 5%.

```{r ss_check-1, exercise=TRUE}
# enter assumed values
p <- 
n <- 

# calculate the margin of error (MOE)
z <- 1.96
z*sqrt(p*(1 - p) / n)
```

```{r ss_check-1-solution}
# enter assumed values
p <- 0.26
n <- 296

# calculate the margin of error (MOE)
z <- 1.96
z*sqrt(p*(1 - p) / n)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should obtain a margin of error of 4.997%, which is very close to the target 5%. Notice that our MOE will always be equal or *smaller* than the target MOE. This is because we rounded the sample size up from 295.65 to 296.

</details>
</br>

## Buffering

Buffering refers to increasing a sample size to allow for drop-out (loss of samples). Some ways that drop-out can occur are through:

- Participants withdrawing consent
- Participants dying or leaving the area
- Samples being lost during transportation
- Samples becoming contaminated
- Samples failing amplification or sequencing
- Data being lost due to computational errors

We cannot completely eliminate the risk of drop-out, but by buffering sample sizes we can be robust to it. If we expect a proportion $d$ of samples to be lost, then the formula for buffered sample size is:

$$
n_{\text{buffered}} = \frac{n_{\text{original}}}{1 - d}
$$

Through consulting with lab technicians and the study team, you estimate that 10% of samples may be lost to drop-out. Complete the following R code to come up with a buffered sample size:

```{r buffer-1, exercise=TRUE}
# enter sample size and estimated drop-out
n <- 296
d <- 0.1

# calculate the buffered sample size
n_buffered <- 

print(n_buffered)
```

```{r buffer-1-solution}
# enter sample size and estimated drop-out
n <- 296
d <- 0.1

# calculate the buffered sample size
n_buffered <- n / (1 - d)

print(n_buffered)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should obtain a buffered sample size of 328.89, which we would round up to 329.

</details>
</br>


## Accounting for positive fraction

So far, we have focused on working out how many *confirmed malaria cases* we need in our study. However, this will be a cross-sectional study with individuals being sampled at random from households within Rutshuru town, and many of these people will test negative for malaria. It would be useful for the study team to know how many individuals they need to *test* as part of this study, which may be considerably higher than the number of confirmed malaria cases.

The NMCP estimates that 25% of the population of Rutshuru will be positive for malaria by RDT. We can use the same buffering formula as before to account for this, but now using the positive fraction ($f$) to inflate our sample size:

$$
n_{\text{test}} = \frac{n_{\text{confirmed}}}{f}
$$

Note that this inflating of the sample size to account for positive fraction is **on top of** buffering for drop-out. We can imagine a chain of events where we can lose participants at each stage; first by testing negative and second by unexpected drop-out. We need to ensure that in the end we still have enough samples remaining, hence we account for both events.

Complete the following R code to work out the number of people that need to be tested to achieve the final target sample size:

```{r fraction-1, exercise=TRUE}
# enter buffered sample size and positive fraction
n <- 329
f <- 

# calculate the testing sample size
n_test <- 

print(n_test)
```

```{r fraction-1-solution}
# enter buffered sample size and positive fraction
n <- 329
f <- 0.25

# calculate the testing sample size
N_test <- n / f

print(n_test)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

You should find that 1316 people need to be tested.

</details>
</br>

### Final recommendation

You have now completed your study design exercise. Your recommendation to the NMCP is as follows:

*Assuming 26% prevalence of dhps K540E mutations from pilot data, a sample size of* **329** *confirmed malaria cases will be needed in order to estimate prevalence to within 5% margin of error. This number is buffered to allow for 10% drop-out.*

*Assuming that malaria prevalence is 25% by RDT in Rutshuru town, this translates to* **1316** *individuals who will need to be tested in the cross-sectional study design.*

## Bonus questions

Our study design is carefully thought through, and based on strong statistical principles. However, it is worth testing how robust these numbers are to changes in our assumptions.

Under the chosen sample size of 296 (after drop-out), what would be your margin of error under the worst case scenario that the true prevalence of the K540E mutation was actually 50%?

```{r bonus-1, exercise=TRUE}

```

```{r bonus-1-solution}
# enter assumed values
p <- 0.5
n <- 296

# calculate the margin of error (MOE)
z <- 1.96
z*sqrt(p*(1 - p) / n)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

The MOE would increase to 5.7% in the most pessimistic scenario. This is not terribly far from the target 5%.

</details>
</br>

We estimated that we will need to test 1316 people in order to achieve this final sample size, based on an assumed 25% prevalence of malaria in Rutshuru town. But what if during the study we found that malaria prevalence was only 15%? What would be your expected final sample size, and what would be your resulting MOE?

```{r bonus-2, exercise=TRUE}

```

```{r bonus-2-solution}
# enter assumed values
p <- 0.26
n_test <- 1316

# work out final sample size assuming 15% prevalence and 10% drop-out
n <- round(n_test * 0.15 * 0.9)
print(n)

# calculate the margin of error (MOE)
z <- 1.96
z*sqrt(p*(1 - p) / n)
```

<details>
<summary style="text-decoration: underline; color: red;">
`r fontawesome::fa("check", fill = "red")`Click to see the answer
</summary>

Assuming 15% prevalence and 10% drop-out, our final sample size of successfully sequenced malaria cases would be around 178, down from 329. This would result in a MOE of 6.4%.

</details>
</br>

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  How could you change your assumed parameters to make the study more robust? What would be the negative impacts of these changes?
  </span>
</div>
