---
title: "Fact sheet: Statistician"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# options(repos = c(
#   plasmogenepi = "https://plasmogenepi.r-universe.dev",
#   CRAN = "https://cloud.r-project.org"
# ))

library(DRpower)
library(learnr)
knitr::opts_chunk$set(warning = FALSE,
                      echo = FALSE)

# needed for longer calculations (eg DRpower) note - in seconds
tutorial_options(exercise.timelimit = 120)

epi <- data.frame(
  region = c("Coastal Region", "Highland Region", "Forest Region", "Urban Region"),
  prevalence = c("12%", "5%", "28%", "8%"),
  incidence_per_1000_per_year = c(40, 15, 90, 25),
  test_positivity_rate = c("18%", "7%", "35%", "10%"))

health_facilities <- data.frame(
  region = c(
    rep("Coastal Region", 20),
    rep("Highland Region", 10),
    rep("Forest Region", 15),
    rep("Urban Region", 20)
  ),
  health_facility = c(
    paste0("HF", 1:20),    # Coastal Region
    paste0("HF", 21:30),   # Highland Region
    paste0("HF", 31:45),   # Forest Region
    paste0("HF", 46:65)    # Urban Region
  ),
  population_served = c(
    # Coastal Region Population Served
    12000, 14000, 16000, 18000, 20000, 22000, 24000, 26000, 28000, 30000,
    32000, 34000, 36000, 38000, 40000, 42000, 44000, 46000, 48000, 50000,
    # Highland Region Population Served
    10000, 12000, 14000, 16000, 18000, 20000, 22000, 24000, 26000, 28000,
    # Forest Region Population Served
    15000, 18000, 21000, 24000, 27000, 30000, 33000, 36000, 39000, 42000,
    45000, 48000, 51000, 54000, 57000,
    # Urban Region Population Served
    20000, 22000, 24000, 26000, 28000, 30000, 32000, 34000, 36000, 38000,
    40000, 42000, 44000, 46000, 48000, 50000, 52000, 54000, 56000, 58000
  ),
  expected_malaria_cases_per_month = c(
    # Coastal Region Expected Cases per Month
    40, 47, 53, 60, 67, 73, 80, 87, 93, 100,
    107, 113, 120, 127, 133, 140, 147, 153, 160, 167,
    # Highland Region Expected Cases per Month
    13, 15, 18, 20, 23, 25, 28, 30, 33, 35,
    # Forest Region Expected Cases per Month
    113, 135, 158, 180, 203, 225, 248, 270, 293, 315,
    338, 360, 383, 405, 428,
    # Urban Region Expected Cases per Month
    42, 46, 50, 54, 58, 63, 67, 71, 75, 79,
    83, 88, 92, 96, 100, 104, 108, 113, 117, 121
  )
)

cost_per_hf <- 5000
cost_per_sample <- 55
cost_transport_coast <- 1000
cost_transport_highland <- 1500
cost_transport_forest <- 2500
cost_transport_urban <- 800
```

## Your role
As the statistician on the team, your job is to help your team make informed decisions on study design considerations, including:

- Decide on the number of clusters (health facilities) to include.
- Buffer for drop-out.
- Determining the necessary sample size and power for primary and secondary endpoints.
- Adjust for ICC. 

## Available data
In order to help you perform these critical calculations, we have pre-loaded all the fact sheets and data that your various team members have access to. 

This includes: 

- `epi`: data on malaria prevalence, incidence and test positivity rate
- `health_facilities`: data on health facilities in each region, the population they serve, and expected malaria cases per month  
- `cost_per_hf`: fixed cost per health facility enrolled (in USD), including training, equipment, and administrative expenses.
- `cost_per_sample`: cost per sample enrolled (in USD), including collection, laboratory testing, consumables, and data management
- `cost_transport_coast`: transport cost per health facility (in USD) in the Coastal region
- `cost_transport_highland`: transport cost per health facility (in USD) in the Highland region
- `cost_transport_forest`: transport cost per health facility (in USD) in the Forest region
- `cost_transport_urban`:  transport cost per health facility (in USD) in the Urban region

We have also pre-loaded the [DRpower R package](https://mrc-ide.github.io/DRpower/index.html) so you can access its functions if you need to!

### Exercise 

*You can use the below R code box to perform your calculations*

```{r code-box, exercise=TRUE}

```