---
title: "Activity 1: Sampling from a population"
output:
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
# options(repos = c(
#   plasmogenepi = "https://plasmogenepi.r-universe.dev",
#   CRAN = "https://cloud.r-project.org"
# ))

library(learnr)
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
# library(DRpower)

knitr::opts_chunk$set(warning = FALSE,
                      echo = FALSE)

# needed for longer calculations (eg DRpower) note - in seconds
tutorial_options(exercise.timelimit = 120)

# FUNCTIONS
sampleFromPopulation <- function(sample_size, census){
  ss <- sample_size
  sample_data <- census %>% sample_n(ss, replace = F) # sample without replacement 
  
  return(sample_data)
}

compareSampleToPopulation <- function(sample, census){
  comparison <- bind_rows(
    census %>% mutate(group = "Population") %>% select(age, sex, malaria_infection, group),
    sample %>% mutate(group = "Sample") %>% select(age, sex, malaria_infection, group))
  
  return(comparison)
}

plotAgeSexDistribution <- function(comparison){
  comparison %>% 
  ggplot(aes(x = age, fill = sex)) +
    geom_histogram(binwidth = 5, color = "white", boundary = 0) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = "Age and sex distribution", x = "Age (years)", y = "Count") +
    theme_minimal() +
    facet_wrap(~group, scales = "free_y")
}

plotInfectedProportion <- function(comparison){
  comparison %>% 
    count(group, malaria_infection) %>% 
    group_by(group) %>% 
    mutate(proportion = n/sum(n)) %>% 
    ungroup() %>% 
    ggplot(aes(x = group, y = proportion, fill = malaria_infection)) +
      geom_bar(stat = "identity") +
      geom_text(aes(label = scales::percent(proportion, accuracy = 0.1)), 
                color = "white",
                position = position_stack(vjust = 0.5),
                size = 4,
                show.legend = F) +
      scale_fill_brewer(palette = "Set1",
                        name = NULL) +
      scale_color_brewer(palette = "Set1") +
      scale_y_continuous(labels = scales::percent) +
      labs(title = "Malaria infection status", x = "", y = "") +
      theme_minimal()
}

checkPrevalenceCI <- function(true_prevalence, CI_lower, CI_upper){
  if(true_prevalence >= CI_lower & true_prevalence <= CI_upper){
    print(paste0("The true prevalence (", round(true_prevalence*100, 2),"%) falls within the 95%CI (", round(CI_lower*100, 2),"- ", round(CI_upper*100, 2), "%)"))
  }
  else {
    print(paste0("The true prevalence (", round(true_prevalence*100, 2),"%) does not fall within the 95%CI (", round(CI_lower*100, 2),"- ", round(CI_upper*100, 2), "%)"))
  }
}

simulate_CI <- function(census, sample_size, true_prevalence) {
  sample_data <- census %>% sample_n(sample_size, replace = F)
  infected_count <- sum(sample_data$malaria_infection == "Infected")
  p_hat <- infected_count / sample_size
  SE <- sqrt((p_hat * (1 - p_hat)) / sample_size)
  CI_lower <- p_hat - Z * SE
  CI_upper <- p_hat + Z * SE
  (true_prevalence >= CI_lower) & (true_prevalence <= CI_upper)
}

plotCISimulationResults <- function(results, n_simulations){
  results_df <- data.frame(result = results) %>% 
                mutate(result = case_when(result == FALSE ~ "Not within 95%CI",
                                          result == TRUE ~ "Within 95%CI"))
  
  results_df %>% 
  count(result) %>% 
  mutate(proportion = n/sum(n)) %>% 
  ggplot(aes(x = "", y = proportion, fill = result)) +
    geom_bar(stat = "identity",
             position = "fill",
             width = 0.5) +
    geom_text(aes(label = scales::percent(proportion, accuracy = 0.1),
                  color = result), 
            position = position_fill(vjust = 0.9), 
            size = 4,
            show.legend = F) +
    scale_fill_manual(values = c("Not within 95%CI" = "grey", "Within 95%CI" = "magenta4"),
                      name = NULL) +
    scale_color_manual(values = c("Not within 95%CI" = "grey30", "Within 95%CI" = "white"),
                      name = NULL) +
    scale_y_continuous(labels = scales::percent) +
    labs(title = paste0("Simulation results (N=", n_simulations, " simulations)"), x = "", y = "") +
    coord_flip() +
    theme_minimal() +
    theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())
}

# PRE-LOADED DATA

# fake census data
# N <- 10000
# 
# set.seed(1234)
# 
# census <- data.frame(
#   id = 1:N,
#   age = sample(0:80, N, replace = TRUE), # not really realistic, eg 'young population' but maybe OK for these purposes?
#   sex = sample(c("Male", "Female"), N, replace = TRUE) #,
#   # malaria_infection = sample(c("Infected", "Not infected"), N, replace = TRUE, prob = c(0.3, 0.7)
# )
# 
# census <- census %>% 
#   # infection probability higher in children
#   mutate(infection_prob = case_when(age >=0 & age <= 14 ~ 0.60,
#                                     age > 14 & age <=44 ~ 0.30,
#                                     age > 44 ~ 0.10,
#                                     TRUE ~ 0.25),
#          malaria_infection = ifelse(runif(n()) <= infection_prob, "Infected", "Not infected")) %>% 
#   select(-infection_prob)
# 
# saveRDS(census, "data/census.rds")

census <- readRDS("data/census.rds")

# GLOBAL PARAMS FOR EXERCISES
true_infected_count <- sum(census$malaria_infection == "Infected")
true_prevalence <- true_infected_count / 10000

# get random sample
# set.seed(5) # to ensure we get the same random sample for our tutorial calculations

# set the sample size
# sample_size <- 200

# sample from the population
# sample_data <- sampleFromPopulation(sample_size, census) 
# saveRDS(sample_data, "tutorials/data/sample_data.rds")

# sample_data <- readRDS("tutorials/data/sample_data.rds")

# Number of infected individuals in the sample
# infected_count <- sum(sample_data$malaria_infection == "Infected")
# 
# p_hat <- infected_count / sample_size
# SE <- sqrt((p_hat * (1 - p_hat)) / sample_size)
Z <- 1.96
# CI_lower <- p_hat - Z * SE
# CI_upper <- p_hat + Z * SE
```

## Introduction

Welcome to Activity 1: **Sampling from a population**.

In this module, we’ll explore how to relate samples to populations, calculate confidence intervals, and understand the impact of sample size on study results. To introduce these statistical concepts, we’ll start with a simple example focused on determining the prevalence of malaria. This approach provides a gentle introduction to thinking about these ideas in the context of study design. These same concepts are directly applicable to malaria molecular surveillance (MMS) studies. For instance, in MMS studies, we might design our research to estimate the prevalence of molecular markers, such as drug resistance markers, rather than the prevalence of malaria infection itself.

### Learning Outcomes

By the end of this tutorial, you will be able to:

-   Define the target population for a study.
-   Explain the difference between a population and a sample.
-   Calculate a 95% confidence interval.
-   Assess how sampling variability impacts our estimate of the population.
-   Understand the effect of sample size on confidence intervals.

## What is my population?

In any study, clearly defining the **target population** is crucial. Let's consider a few examples of made-up studies.

```{r quiz-study-pop}
quiz(caption = "QUIZ - Target population",
  question_checkbox("We want to estimate the prevalence of malaria infection in a community of 10,000 people. Which of the following is the target population?",
           allow_retry = TRUE,
           answer("All residents", correct = TRUE),
           answer("Only women or only men", correct = FALSE, message = "If we want to estimate the prevalence in the *community* we don't want to focus only on women or men. <br>"),
           answer("Only children under 5 years old", correct = FALSE, message = "If we want to estimate the prevalence in the *community* we don't want to focus only on children. <br>"),
           answer("Adults over 60 years old", correct = FALSE, message = "If we want to estimate the prevalence in the *community* we don't want to focus on adults. <br>"), 
           correct = "That is correct! Our target population should be a sample of all the community residents, not a specific demographic or group.", 
           incorrect = "If we want to estimate the prevalence in the *community* we don't want to focus on a specific demographic or group. <br>"
  )
)
```

## Relating the sample to the population

Sampling the entire population is often impractical, whether due to cost constraints or the challenges of reaching everyone. Fortunately, we can design our study to ensure that the sample we collect is a **representative subset** of the population.

When designing epidemiological studies, various data sources can offer valuable insights into the population. Examples include Demographic and Health Surveys (DHS) or population censuses, which provide comprehensive information to help guide study design.

### Using the population census

Let's go back to our made-up study.

```{r}
# total population in census
N <- 10000

# infected population
infected_pop <- census %>% filter(malaria_infection == "Infected")
```

We now have access to a census of the entire population of the village (N = `r as.integer(N)`) with information on each resident's age and sex. For purposes of this tutorial, imagine there are `r census %>% filter(malaria_infection == "Infected") %>% nrow()` people in the village with malaria infections and we know the true infection status of every individual. This will help us understand how our sample relates to the entire population.

This is the information we have in the census (here we show the first 6 residents in the census):

```{r}
#kable(head(census))
kable(head(census)) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("striped", "hover"))
```

### Population demographics
Below we can see a breakdown of malaria-infected individuals in this population.

```{r}
census %>% 
  count(malaria_infection) %>% 
  mutate(proportion = n/sum(n)) %>% 
  ggplot(aes(x = malaria_infection, y = n, fill = malaria_infection)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = scales::percent(proportion, accuracy = 0.1),
                  color = malaria_infection), 
            nudge_y = 200,
            size = 4,
            show.legend = F) +
    scale_fill_brewer(palette = "Set1") +
    scale_color_brewer(palette = "Set1") +
    labs(title = "Malaria infection status", x = "", y = "Count") +
    guides(fill = F) +
    theme_minimal()
```

Let's look at the age and sex distribution.

```{r}
infected_pop %>% 
  ggplot(aes(x = age, fill = sex)) +
    geom_histogram(binwidth = 5, color = "white", boundary = 0) +
    scale_fill_brewer(palette = "Set2") +
    labs(title = "Malaria-infected population", x = "Age (years)", y = "Count") +
    theme_minimal()
```

### Sampling from the population

Suppose we have resources to determine the infection status of 200 individuals. Let's randomly sample 200 individuals from the population and see how they compare with the overall population. We can explore demographic differences by plotting the age and sex distribution of the sample compared with the population.

Below we use the function `sampleFromPopulation()` and we specify what our sample size is and from where we want to sample (in our case from the census data). We can use the `plotAgeSexDistribution()` and `plotInfectedProportion()` functions to visualize our results. Click on "Run Code". 

```{r sampling-1, exercise=TRUE}
sample <- sampleFromPopulation(sample_size = 200, census)
comparison <- compareSampleToPopulation(sample, census)

plotAgeSexDistribution(comparison)

plotInfectedProportion(comparison)
```

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  Do we see a similar age and sex distribution? What about the proportion of infected and not infected individuals in the population?
  </span>
</div>

### Sampling many times from the population 

Now run this a few times by clicking "Run Code" to see how it changes with every random sample.

```{r sampling-2, exercise=TRUE}
sample <- sampleFromPopulation(sample_size = 200, census)
comparison <- compareSampleToPopulation(sample, census)

plotAgeSexDistribution(comparison)
plotInfectedProportion(comparison)
```

```{r echo=F, eval=F}
# THIS IS A SANITY CHECK OF THE DISTRIBUTION - don't think we need this in the actual tutorial?

n_sim <- 1000
# Vector to store sample proportions
sample_proportions <- numeric(n_sim)

for (i in 1:n_sim) {
  sample_data <- census %>% sample_n(sample_size, replace = F)
  
  # Calculate sample proportion
  p_hat <- mean(sample_data$malaria_infection == "Infected")
  
  # Store the sample proportion
  sample_proportions[i] <- p_hat
}

# Calculate the mean and standard deviation of sample proportions
mean_p_hat <- mean(sample_proportions)
sd_p_hat <- sd(sample_proportions)

# Display the results
cat("Mean of sample proportions:", mean_p_hat, "\n")
cat("Standard deviation of sample proportions:", sd_p_hat, "\n")

# Plot histogram of sample proportions
hist(sample_proportions, breaks = 30, main = "Distribution of Sample Proportions", xlab = "Sample Proportion of Infected Individuals")
```


```{r quiz-sampling}
quiz(caption = "QUIZ - Sampling from the population",
  question_radio("Was the sample always the same?",
           answer("Yes", correct = FALSE),
           answer("No", correct = TRUE),
           allow_retry = TRUE,
           correct = "That is correct! Our sample changes every time we randomly draw a sample from a population.",
           incorrect = "Are you sure? Try running the code above a few more times. Is it always reflective of the population in terms of age and sex distribution and/or with respect to the proportion of people infected?"
  ),
  question_radio("Does our sample of 200 always reflect the population?",
           answer("Yes", correct = FALSE),
           answer("No", correct = TRUE),
           allow_retry = TRUE,
           correct = "That is correct! When we have small sample sizes this can lead to high uncertainty and more variability thus our sample may not be representative of the population.",
           incorrect = "Are you sure? Try running the code above a few more times. You could even replace 200 with 50 to see how this impacts the results. Is it always reflective of the population in terms of age and sex distribution and/or with respect to the proportion of people infected? Smaller sample sizes are more susceptible to sampling variability. With a limited number of individuals, the likelihood of the sample deviating from the population characteristics increases."
  )
)
```

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  Smaller sample sizes are more susceptible to sampling variability. With a smaller number of individuals there is an increased chance the sample will deviate from the population characteristics. Think about how this may impact your results. 
  </span>
</div>

### ✨ BONUS QUESTION ✨

You will have noticed from our exploration above that the sample differs from the population and it doesn't always have the same age and sex distribution.

```{r quiz-sampling-bias}
quiz(caption = "QUIZ - Sampling bias",
  question_radio("Does this mean our sampling approach is biased?",
           answer("Yes", correct = FALSE),
           answer("No", correct = TRUE),
           allow_retry = TRUE,
           correct = "That is correct! The fact that the sample differs from the population does not necessarily mean our sampling approach is biased. Bias refers to a *systematic error* that consistently favors certain outcomes. However, because we are randomly sampling 200 individuals with *equal chance of being selected* our sampling approach is not biased. This can occur even if the characteristics of our sample (e.g., age and sex distribution) may not perfectly match those of the population.",
           incorrect = "That is not correct! Although our sample was not always reflective of the population (for example, the characteristics like age and sex distribution did not perfectly match), our *sampling* method was unbiased because we randomly sampled 200 individuals and each person had an equal chance of being selected. Bias refers to a *systematic error* that consistently favors certain outcomes, which was not the case here."
  )
)
```

## Estimating prevalence in our sample and calculating the 95% confidence interval

Our next topic focuses on calculating the 95% confidence interval (CI) on the prevalence. The 95% CI provides an interval with defined lower and upper bounds, representing the range within which we are 95% confident the true population prevalence lies. In other words, if we were to repeat the sampling process many times and calculate a confidence interval for each sample, approximately 95 out of 100 of these intervals would contain the true population prevalence.

### Estimating prevalence in our sample
For this exercise we have already pre-calculated some useful parameters:

- Defined `sample_size` to be 200
- Defined `infected_count` to be the number of infected individuals in our sample (in our example it is 69 individuals)
- Defined the function `sampleFromPopulation()` to select 200 individuals at random from our census

```{r sample-seed-params}
set.seed(5) # to ensure we get the same random sample for our tutorial calculations

# set the sample size
sample_size <- 200

# sample from the population
sample_data <- sampleFromPopulation(sample_size, census) 

# Count number of infected individuals in the sample
infected_count <- sum(sample_data$malaria_infection == "Infected")

# p_hat
p_hat <- infected_count/sample_size

# Z
Z <- 1.96

# SE
SE <- sqrt((p_hat * (1 - p_hat)) / sample_size)

# CI
CI_lower <- p_hat - Z * SE
CI_upper <- p_hat + Z * SE
```

### What is the estimated prevalence of malaria in our sample?
We can calculate this by dividing the number of individuals infected with malaria by our sample size. 

Try coding it yourself, or click on the solution. *Note: In R when we want to divide two things we can use `/`.*

```{r est-prev, exercise = TRUE, exercise.setup = "sample-seed-params"}

```

```{r est-prev-solution}

# Calculate from the raw values
69 / 200

# Or you can use the stored variables:
infected_count / sample_size
```

<details>

<summary style="text-decoration: underline; color: red;">

`r fontawesome::fa("check", fill = "red")`Click to see the answer

</summary>
Our estimated prevalence is 0.345 or 34.5%.

### Calculating the 95% CI

Next, we need to calculate the 95% CI around our point estimate.

Below is the formula for the **Wald confidence interval:**

$$
CI = \hat{p} \pm z_{1 - \alpha/2}\sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
$$
Where:

- $\hat{p}$ = sample proportion (i.e. estimated prevalence)
- ${n}$ = sample size
- $\ z_{1 - \alpha/2}$ = critical value of the normal distribution at significance level $\alpha$ (two sided)

We will now go through this formula step-by-step!

#### 1. Defining our sample proportion, $\hat{p}$

The sample proportion refers to the proportion of infected individuals in our sample. We calculated this above by dividing the number of infected individuals in the sample by the total sample size.

Complete the code below to calculate this value and record it as `p_hat`. 

```{r phat, exercise=TRUE, exercise.setup = "sample-seed-params"}

p_hat <-
```

```{r phat-solution}

# Sample proportion
p_hat <- infected_count / sample_size
p_hat
```

##### 2. Sample size, $n$

Above, we defined our sample size to be n=200. We already recorded this in the variable `sample_size`, so you don't need to do anything extra.

##### 3. Calculating the standard error

We can calculate the standard error using `p_hat` and `sample_size`. We can calculate the standard error using:

$$
SE = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}
$$
Complete the code below to calculate standard error and store it in the variable `SE`:

*Note: In R the function to take the square root is `sqrt()` and when we want to multiply two things we can use `*`.*

```{r SE, exercise=TRUE, exercise.setup = "sample-seed-params"}

SE <- 
```

```{r SE-solution}

# Standard error
SE <- sqrt((p_hat * (1 - p_hat)) / sample_size)
SE
```

<details>

<summary style="text-decoration: underline; color: red;">

`r fontawesome::fa("check", fill = "red")`Click to see the answer

</summary>
Our standard error is 0.034.

##### 4. Calculating the confidence interval

Now that we know our standard error, we can calculate the lower and upper bounds of our 95% confidence interval (CI). To do this, we multiply the standard error by  $z_{1-\alpha/2}$ , which is approximately 1.96 for a 95% CI. This factor is derived from the normal distribution, where 95% of the probability lies within 1.96 standard deviations of the mean.

Complete the code below to calculate the lower bound from `z_alpha` and your stored value of `SE`. Look back at the formula for the Wald interval above if you're stuck.
```{r ci-lower, exercise=TRUE, exercise.setup = "sample-seed-params"}

# Critical value for 95% confidence
z_alpha <- 1.96

# Lower bound

```

```{r ci-lower-solution}

# Critical value for 95% confidence
z_alpha <- 1.96

# Lower bound
p_hat - z_alpha * SE
```

<details>

<summary style="text-decoration: underline; color: red;">

`r fontawesome::fa("check", fill = "red")`Click to see the answer

</summary> 
Our lower bound is 0.279 or 27.9%. 

</details>

Now let's calculate the upper bound. Remember now we need to *add* instead of subtract. 
```{r ci-upper, exercise=TRUE, exercise.setup = "sample-seed-params"}

# Critical value for 95% confidence
z_alpha <- 1.96

# Upper bound

```

```{r ci-upper-solution}

# Critical value for 95% confidence
z_alpha <- 1.96

# Upper bound
p_hat + z_alpha * SE
```

<details>

<summary style="text-decoration: underline; color: red;">

`r fontawesome::fa("check", fill = "red")`Click to see the answer

</summary>
Our upper bound is 0.411 or 41.1%. 
</details>

##### 5. Putting it all together

Let's put all the steps above together! 
```{r ci-calculation, exercise=TRUE}
# input observed data
infected_count <- 69
sample_size <- 200

# calculate the sample proportion and standard error
p_hat <- infected_count / sample_size
SE <- sqrt((p_hat * (1 - p_hat)) / sample_size)

# calculate lower and upper 95% bounds
z_alpha <- 1.96
CI_lower <- p_hat - z_alpha * SE
CI_upper <- p_hat + z_alpha * SE

# Print our values
p_hat
CI_lower
CI_upper
```

So, putting it all together, our estimated prevalence is 34.5% and our 95% CI is 27.9% to 41.1%.

## Comparing to the true prevalence

You may remember from the census data that **26.78%** of our population was infected with malaria. This is the *true* prevalence.

```{r quiz-ci-prev}
quiz(caption = "QUIZ - 95% confidence interval",
     question("Our 95% CI goes from 27.9% to 41.1%. Is our true prevalence within the 95% CI?",
              allow_retry = TRUE,
              answer("Yes", correct = F, message = "The true prevalence (26.78%) is not within the 95%CI of 27.9-41.1%"),
              answer("No", correct = T, "The true prevalence (26.78%) is not within the 95%CI of 27.9-41.1%. In the next exercise we are going to run some simulations to explore this further.")
     ))
```

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  Do you always expect the true prevalence to fall within the 95%CI?
  </span>
</div>

## How often does true prevalence fall within the 95%CI?

We explored above what would happen if we randomly sampled 200 individuals. In our example, the true prevalence didn't fall within the 95% CI. But this was just one example. Now we want to see what happens if we repeat this sampling *many* times. Let's now explore by running a simulation where we sample 1000 times and we will count how many times our true prevalence is within the 95%CI.

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  Before you run the below code, think about the intuition behind this - how often do you expect the true prevalence to be within the 95%CI? 
  </span>
</div>

Now run the code and see if you were correct!

```{r sim-ci, exercise=TRUE}
n_simulations <- 1000
sample_size <- 200

results <- replicate(n_simulations, simulate_CI(census, sample_size, true_prevalence))

plotCISimulationResults(results, n_simulations)
```


```{r quiz-sim-ci}
quiz(
  caption = "QUIZ - Confidence Intervals and Simulation",

  question("In the context of confidence intervals, what does a 95% confidence level mean?",
           allow_retry = TRUE,
    answer("It means that 95% of the time, the sample estimate will be equal to the true population parameter.", correct = FALSE),
    answer("It means that 95% of the time the true population parameter will lie within this interval.", correct = TRUE),
    answer("It means that 95% of the data points lie within the confidence interval.", correct = FALSE),
    answer("If we take 100 samples, exactly 95 of them will have intervals containing the true parameter.", correct = FALSE),
    correct = "Correct! A 95% confidence level means we are 95% confident that the true population parameter lies within the interval calculated from our sample.",
    incorrect = "Not quite. Remember the simulation exercise we undertook above."
  ),

  question("After running the simulation of 1,000 samples, approximately how many of the confidence intervals would you expect to contain the true population prevalence?",
           allow_retry = TRUE,
    answer("About 50 of them.", correct = FALSE),
    answer("About 950 of them.", correct = TRUE),
    answer("All 1,000 of them.", correct = FALSE),
    answer("None of them.", correct = FALSE),
    correct = "That's right! With a 95% confidence interval, we'd expect about 950 out of 1,000 intervals to contain the true prevalence.",
    incorrect = "Consider the definition of a 95% confidence interval—it suggests that approximately 95% of intervals contain the true parameter."
  ),

  question("If the true prevalence falls outside the confidence interval calculated from our sample, does it mean our sampling method is biased?",
           allow_retry = TRUE,
    answer("Yes, because the confidence interval should always contain the true parameter.", correct = FALSE),
    answer("No, because even with a correct method, there's still a chance the confidence interval won't contain the true parameter.", correct = TRUE),
    answer("Yes, because our sample size was too small.", correct = FALSE),
    answer("No, because the confidence interval is only a rough estimate.", correct = FALSE),
    correct = "Exactly! Even with an unbiased sampling method, there's a 5% chance the true parameter won't be within the 95% confidence interval.",
    incorrect = "Not quite. Remember that confidence intervals are calculated with a certain level of confidence, and there's always a chance the true parameter lies outside the interval."
  )
)
```

After running the simulation, you should notice that the true prevalence falls within the 95% confidence intervals in approximately **95% of the simulations**. This outcome aligns with the definition of a 95% confidence interval: if we were to repeat the sampling process many times, we would expect the true parameter to lie within the calculated confidence interval about 95 out of 100 times.

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  This simulation demonstrates the concept of confidence level in statistical inference. It shows that the method we use to calculate confidence intervals is reliable in the long run. However, in any single sample (like the one we initially took), there's still a chance (about 5%) that the true prevalence will not be captured within the interval. This is why it's important to interpret confidence intervals correctly and understand that they provide a measure of the uncertainty associated with our estimates.
  </span>
</div>

### ✨ BONUS QUESTIONS ✨

*Note: This section is optional and requires more coding than the previous exercise*

Let's repeat this exercise for a sample size of 500. Try coding it yourself from scratch using the functions that we used above. Click on the solution if you get stuck! 

```{r sampling-500, exercise=TRUE}
sample_size <- 500
```

```{r sampling-500-solution}
sample_size <- 500
sample_data <- sampleFromPopulation(sample_size, census)
comparison <- compareSampleToPopulation(sample_data, census)

plotAgeSexDistribution(comparison)
plotInfectedProportion(comparison)

infected_count <- sum(sample_data$malaria_infection == "Infected")

p_hat <- infected_count / sample_size
SE <- sqrt((p_hat * (1 - p_hat)) / sample_size)
CI_lower <- p_hat - Z * SE
CI_upper <- p_hat + Z * SE

checkPrevalenceCI(true_prevalence, CI_lower, CI_upper)

n_simulations <- 1000
results <- replicate(n_simulations, simulate_CI(census, sample_size, true_prevalence))
plotCISimulationResults(results, n_simulations)
```

<div style="padding: 10px; border-radius: 5px; background-color: #fef3e7;">
  <span style="font-size: 1.2em; color: #d19554;">
    <i class="fas fa-comment"></i> Reflection:
  </span> 

  <span style="color:#d19554;"> 
  Is there a greater chance of the true prevalence being within the 95% CI with this larger sample? If not, why not?
  </span>
</div>
